---
title: "English Learners Status in Oregon"
author: Nafisul Huq
output: 
  flexdashboard::flex_dashboard:
    theme:
      bg: "white"
      fg: "black" 
      primary: "#00796B"
    orientation: rows
    vertical_layout: scroll
---

```{r}
library(readr)
library(tidyverse)
library(purrr)
library(sf)
library(tigris)
library(leaflet) 
library(tmap)
library(flexdashboard)
library(ggplot2)
library(ggExtra)
library(grid)
library(plotly)
library(leaflet)
library(viridis)
```

Overview {data-orientation=rows}
=======================================
Sidebar Title {.sidebar data-width=450}
---------------------------------------


### **Understanding English Learner Enrollment in Oregon (2009–2014)**


**English Language Learners (ELL) are students who require additional support in acquiring English proficiency while navigating their academic curriculum. Understanding where EL students are concentrated, how their numbers have changed over time, and what factors influence their enrollment is essential for policymakers and educators working to provide equitable educational opportunities.**

**This dashboard explores trends in EL student enrollment across Oregon's school districts from 2009 to 2014. It examines the relationship between EL populations and key demographic factors, including urban vs. rural district differences and the link between socioeconomic status (free lunch eligibility) and EL enrollment.**
<br>
</br>



### Research Questions
**This analysis focuses on two key questions:**
<br>
</br>
**1. How have EL student enrollment trends changed over time in Oregon?**
<br>
</br>
**2. How do factors like economic disadvantage and district location (urban vs. rural) influence EL enrollment?**

<br>
</br>



Row {data-height=800}
-----------------------------------------




### **Trends in English Learner Enrollment in Oregon (2009–2013)**


This section presents a longitudinal analysis of the total number of English Learner (EL) students enrolled across school districts in Oregon. Monitoring these trends helps stakeholders understand changes in EL populations over time, potentially informing resource allocation, support services, and educational policy decisions.

```{r, include=FALSE}



data_oregon <- read_csv("C:/Users/Huq/Desktop/EDLD640-Capstone/Capstone640/data_oregon.csv")
```


```{r, include=FALSE}
data_oregon <- data_oregon %>%
  mutate(leaname = case_when(
    leaname == "GREATER ALBANY PUBLIC SD 8J"  ~ "GREATER ALBANY SD 8J",
    leaname == "ATHENA-WESTON SD 29RJ"  ~ "ATHENA-WESTON SD 29J",
    leaname == "MT ANGEL SD 91"  ~ "MOUNT ANGEL SD 91",
    leaname == "UKIAH SD 80R"  ~ "UKIAH SD 80",
    leaname == "UMATILLA SD 6R"  ~ " UMATILLA SD 6",
    leaname == "WEST LINN-WILSONVILLE SD 3J"  ~ "WEST LINN SD 3J",
    leaname == "ST PAUL SD 45"  ~ "ST. PAUL SD 45",
    leaname == "ST HELENS SD 502"  ~ "ST. HELENS SD 502",
    leaname == "BROOKINGS-HARBOR SD 17C"  ~ "BROOKINGS-HARBOR SD 17",
    leaname == "BEND-LAPINE ADMINISTRATIVE SD 1"  ~ "BEND-LA PINE ADMINISTRATIVE SD 1",
    leaname == "BLACHLY SD 90"  ~ "BLACHLY SD 090",
    leaname == "PINE EAGLE SD 61"  ~ "PINE-EAGLE SD 61",
    leaname == "PORT ORFORD-LANGLOIS SD 2CJ"  ~ "PORT ORFORD-LANGLOIS SD 2J",
    leaname == "ONTARIO SD 8C"  ~ "ONTARIO SD 8",
    leaname == "MEDFORD SD 549C"  ~ "MEDFORD SD 549",
    leaname == "MILTON-FREEWATER UNIFIED SD 7"  ~ "MILTON-FREEWATER SD 7",
    leaname == "HOOD RIVER COUNTY SD"  ~ "HOOD RIVER COUNTY SD 1",
    leaname == "GRESHAM-BARLOW SD 10J"  ~ "GRESHAM-BARLOW SD 1J",
    leaname == "LAKE COUNTY SD 7"  ~ "LAKEVIEW SD 7",
    leaname == "SOUTH LANE SD 45J3"  ~ "SOUTH LANE SD 45J",
    leaname == "YAMHILL CARLTON SD 1"  ~ "YAMHILL-CARLTON SD 1",
    leaname == "NORTH WASCO COUNTY SD 21"  ~ "NORTH WASCO SD 21",
    leaname == "IONE SD R2"  ~ "IONE SD 2",
    leaname == "SHERMAN COUNTY SD"  ~ "SHERMAN SD 1",
    leaname == "THREE RIVERS/JOSEPHINE COUNTY SD"  ~ "THREE RIVERS SD",
    #leaname == " UMATILLA SD 6"  ~ " UMATILLA SD 6",
    leaname == "DOUGLAS COUNTY SD 4" ~ "HARNEY COUNTY SD",
    TRUE ~ leaname
  ))

```

```{r, include=FALSE}
oregon_ell <- data_oregon %>%
  group_by(leaid,year) %>%
  summarise(ell  = first(ell), .groups = "drop")


oregon_ell_distribution <- oregon_ell %>%
  group_by(year) %>%
  summarise(total_ell = sum(ell, na.rm = TRUE)) %>%
  rename(c(Year = year, 'Total EL Student' = total_ell))


```



```{r, include=FALSE}
p1 <- ggplot(oregon_ell_distribution, aes(x = Year, y = `Total EL Student`)) +
  geom_line(color = "#00796B", size = 1.2) +  # Teal line for consistency
  geom_point(shape=21, color = "black", fill= "#D81B60", size = 4, stroke=1) +   # Magenta for contrast & visibility
  labs(title = "EL Student Distribution in Oregon by Year",
       x = "Year",
       y = "Total EL Students") +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "#00796B", size = 14, face = "bold"),  # Matching title color
    axis.text = element_text(color = "black"),  # Keeping text readable
    axis.title = element_text(color = "#00796B", face = "bold")  # Matching axis labels
  )

ggplotly(p1)

```

```{r}
p1 <- ggplot(oregon_ell_distribution, aes(x = Year, y = `Total EL Student`)) +
  geom_line(color = "#009688", size = 1.2) +  # Soft teal (lighter than #00796B)
  geom_point(fill = "#FFC107", size = 4, color = "black", stroke=0.5, shape=21) +   # Soft golden yellow for contrast
  labs(title = "EL Student Distribution in Oregon by Year",
       x = "Year",
       y = "Total EL Students") +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "#009688", size = 14, face = "bold"),  # Soft teal title
    axis.text = element_text(color = "black"),  # Keeping text readable
    axis.title = element_text(color = "#009688", face = "bold")  # Matching axis labels
  )

ggplotly(p1)  %>% layout(height = 500)

```


Row {data-height=300}
-----------------------------------------
### Data Description

This dashboard leverages publicly available data from the Stanford Education Data Archive (SEDA), provided by the [Stanford Education Opportunity Project](https://edopportunity.org) . The dataset includes school district-level demographic and performance indicators, allowing for a detailed examination of English Learner (EL) student enrollment trends in Oregon from 2009 to 2014.
<br>
</br>
Through interactive visualizations—including line graphs, bar charts, and spatial maps—this dashboard tracks changes in EL enrollment across urban and rural school districts, as well as its correlation with socioeconomic factors such as free lunch eligibility. These insights provide valuable context for educators, policymakers, and researchers working to understand how demographic disparities impact EL student distribution and educational opportunities across Oregon.
<br>
</br>



#  Area type 

**How Do Urban and Rural Districts Compare?**
<br>
</br>
This section compares EL student enrollment in urban vs. rural school districts, showing the differences in student populations.
<br>
</br>

```{r, include=FALSE}
oregon_urban <- data_oregon %>%
  group_by(leaid,year) %>%
  summarise(urban  = first(urban), .groups = "drop")  

```

```{r,include=FALSE}


# Join the two datasets by leaid and year
oregon_ell_urban <- oregon_ell %>%
  full_join(oregon_urban, by = c("leaid", "year"))



```

```{r,include=FALSE}


# Aggregate EL student count by year and urban/rural classification
oregon_ell_urban_summary <- oregon_ell_urban %>%
  group_by(year, urban) %>%
  summarise(total_ell = sum(ell, na.rm = TRUE), .groups = "drop")



# Convert urban to a factor with correct labels and drop unused levels
oregon_ell_urban_summary$urban <- factor(oregon_ell_urban_summary$urban, 
                                         levels = c(0, 1), 
                                         labels = c("Rural", "Urban"))

# Ensure no extra levels are present
oregon_ell_urban_summary <- na.omit(oregon_ell_urban_summary)

oregon_ell_urban_summary <-oregon_ell_urban_summary %>%
  rename(c( Year              = year, 
            Student = total_ell,
            Area             = urban))
```


```{r, echo=FALSE}
# Plot the data
p2 <- ggplot(oregon_ell_urban_summary, aes(x = Year, y = Student, fill = Area)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Rural" = "#1f77b4", "Urban" = "#ff7f0e")) +
  labs(title = "EL Student Distribution by Urban & Rural Areas",
       x = "Year",
       y = "Total EL Students",
       fill = "Area Type") +
  theme_minimal()

ggplotly(p2)

```

<br>
</br>
The bar chart consistently shows that urban districts have a significantly higher number of EL students compared to rural districts. Although both urban and rural EL enrollments fluctuate slightly over time, the urban EL student population remains dominant throughout. This suggests that EL students are more concentrated in urban areas, likely due to larger immigrant populations and better access to language support programs in city school districts.
<br>
</br>


```{r}
p3 <- ggplot(oregon_ell_urban_summary, aes(x = Year, y = Student, color = Area, group = Area)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Trend of EL Students in Urban vs. Rural Areas",
       x = "Year",
       y = "Total EL Students",
       color = "Urban/Rural") +
  theme_minimal()

ggplotly(p3)

```
<br>
</br>
This line graph confirms that urban districts maintain a steady EL student population, whereas rural districts show more variation. The EL student population in urban areas remains relatively stable, while rural districts experience a slight decline over time. This could be due to demographic shifts, with families moving to urban centers for better job opportunities and educational resources.
<br>
</br>

Row {data-height=400}
-----------------------------------------

### Tab
> This visualization offers 


#  Socioeconomic Factors & EL Students

Economic Disadvantage & EL Enrollment

This section examines how free lunch eligibility (a proxy for economic status) relates to EL student distribution.

```{r, include=FALSE}
oregon_frl <- data_oregon %>%
  group_by(leaid,year) %>%
  summarise(perfrl  = first(perfrl)*100, .groups = "drop")


# Join the two datasets by leaid and year
oregon_frl <- oregon_ell %>%
  full_join(oregon_frl, by = c("leaid", "year"))


```

```{r}
oregon_frl %>%
  group_by(year) %>%
  summarise(avg_perfrel = mean(perfrl, na.rm = TRUE),
            avg_ell = mean(ell, na.rm = TRUE)) %>%
  pivot_longer(cols = c(avg_perfrel, avg_ell), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = year, y = Value, color = Variable)) +
  geom_line(size = 1.2) +
  labs(title = "Trend of Free Lunch Eligibility and ELL Students Over Time",
       x = "Year",
       y = "Percentage / Count",
       color = "Variable") +
  theme_minimal()

```

The graph presents two overlapping trends:

Free lunch eligibility (%) remains relatively high across all years, indicating that a large proportion of students in Oregon come from low-income backgrounds.
EL student numbers, on the other hand, show a decline over time, reinforcing the trend seen in the first graph.
This suggests that although economic hardship persists, the EL student population is shrinking. One possible explanation is that some EL students may be transitioning out of EL classification faster, while overall immigration trends could also be changing.
```{r}
oregon_frl <- oregon_frl %>%
  mutate(perfrel_category = cut(perfrl, 
                                breaks = c(0, 25, 50, 75, 100), 
                                labels = c("Low (0-25%)", "Medium (26-50%)", 
                                           "High (51-75%)", "Very High (76-100%)"),
                                include.lowest = TRUE))

ggplot(oregon_frl, aes(x = perfrel_category, y = ell, fill = perfrel_category)) +
  stat_summary(fun = mean, geom = "bar", color = "black") +
  labs(title = "Average ELL Students by Free Lunch Eligibility Categories",
       x = "Free Lunch Eligibility (%)",
       y = "Average ELL Students") +
  scale_fill_viridis_d() +
  theme_minimal()

```


```{r, include=FALSE}


oregon_race <- data_oregon %>%
  group_by(leaid,leaname,year) %>%
  summarise( white           = sum(wht), 
             black           = sum(blk),
             hispanic        = sum(hsp),
             asian           = sum(asn),
             native_american = sum(ind),
             total_enrolment = sum(totenrl),
             .groups         = "drop") 


# Joining EL data with race


oregon_race <- left_join(oregon_race, oregon_ell, by = c('leaid','year'))
districts <- st_read("C:/Users/Huq/Desktop/EDLD640-Capstone/Capstone640/gz_2010_41_970_00_500k-20250311T050443Z-001/gz_2010_41_970_00_500k/gz_2010_41_970_00_500k.shp")

districts$NAME <- toupper(districts$NAME)

districts <- districts %>%
  rename(leaname = NAME ) 
  
  
districts1 <- districts %>%
  select(leaname,CENSUSAREA,geometry)

districts1$leaname <- gsub("SCHOOL DISTRICT", "SD", districts1$leaname, ignore.case = TRUE)


districts1 <- districts1 %>%
  mutate(leaname = case_when(
    leaname == "REMAINDER OF OREGON"  ~ "HARNEY COUNTY SD" ,
    leaname == "UMATILLA SD 6" ~ " UMATILLA SD 6" ,
    leaname == "SD NOT DEFINED"  ~ "PINE CREEK SD 5" ,
    leaname == "OREGON TRAIL SD 46"  ~ "SUNTEX SD 10" ,
    TRUE ~ leaname
  ))


oregon_race <- oregon_race %>%
  mutate(leaname = case_when(
    leaname == "JUNTURA SD 12"  ~ "MCDERMITT SD 51" ,
    leaname == "FRENCHGLEN SD 16" ~ "SILVER FALLS SD 4J" ,
    leaname == "DIAMOND SD 7"  ~ "DAYS CREEK SD 15" ,
    leaname == "DREWSEY SD 13" ~ "ROSEBURG SD 4" ,
    TRUE ~ leaname
  ))
# Create a color palette with bins
#pal <- colorBin(palette = "YlOrRd",  # Yellow-Orange-Red scale
               # domain = oregon_race09$ell, bins = bins)



```



```{r, eval=FALSE}

#comm <- intersect(oregon_race09$leaname,districts1$leaname)
oregon_race9 <- oregon_race %>%
  filter(year == 2009)

oregon_race10 <- oregon_race %>%
  filter(year == 2010)

oregon_race11 <- oregon_race %>%
  filter(year == 2011)

oregon_race12 <- oregon_race %>%
  filter(year == 2012)

oregon_race13 <- oregon_race %>%
  filter(year == 2013)

oregon_race14 <- oregon_race %>%
  filter(year == 2014)





# Get district names from both datasets
districts_shape <- unique(districts1$leaname)
districts_race09 <- unique(oregon_race_filtered$leaname)

# Identify uncommon districts
unmatched_in_race09 <- setdiff(districts_shape, districts_race09)
unmatched_in_shape <- setdiff(districts_race09, districts_shape)

# Display the names clearly
cat("Districts in shapefile not in race09 data:\n")
print(unmatched_in_race09)

cat("\nDistricts in race09 data not in shapefile:\n")
print(unmatched_in_shape)





```





Row {data-height=550}
-----------------------------------------
###  Description

Understanding the Economic Context of English Learners in Oregon
This visualization examines the relationship between free lunch eligibility and the number of English Language Learner (ELL) students across school districts in Oregon. Free lunch eligibility serves as a proxy for economic disadvantage, providing insights into how socioeconomic factors impact ELL populations.

<br>
</br>

**Higher ELL Concentration in Economically Disadvantaged Districts**

Districts where 76-100% of students qualify for free lunch have the highest number of ELL students.
This trend suggests that ELL populations are more concentrated in lower-income communities.

<br>
</br>

**Wealthier Districts Have Fewer ELL Students**

School districts with 0-25% free lunch eligibility have significantly fewer ELL students.
This pattern highlights a demographic divide where ELL students are less prevalent in wealthier areas.

<br>
</br>

**Middle-Tier Districts Show Varied ELL Counts**

Districts in the 26-50% and 51-75% free lunch categories show moderate ELL numbers, reinforcing that ELL students are often present in economically mixed communities.



This suggests a strong correlation between economic disadvantage and higher EL student enrollment, reinforcing the idea that many EL students come from economically disadvantaged backgrounds. It also highlights the need for additional support services in lower-income districts where EL populations are concentrated.










Column
-----------------------------------------
### Tab


Column {data-height=500}
-----------------------------------------

### Tab













Map {data-orientation=rows}
==============================

```{r, echo=FALSE}
options(tigris_use_cache = TRUE)

# Define Oregon state boundary
oregon_boundary <- states(cb = TRUE) %>%
  filter(STUSPS == "OR")

# Filter race data for years 2009 to 2014
oregon_race_filtered <- oregon_race %>%
  filter(year >= 2009 & year <= 2014)

# Merge with district shapefile
oregon_race_merged <- full_join(districts1, oregon_race_filtered, by = "leaname")

# Define color palette based on ELL students
pal <- colorNumeric(palette = "viridis", domain = oregon_race_merged$ell, reverse = TRUE)

# Create base leaflet map
map <- leaflet(options = leafletOptions(zoomControl = FALSE, dragging = FALSE, attributionControl = FALSE)) %>%
  addProviderTiles("CartoDB.Positron") %>%  # White background
  addPolygons(
    data = oregon_boundary,
    color = "black", weight = 2, fillOpacity = 0,
    highlightOptions = highlightOptions(weight = 3, color = "black", bringToFront = TRUE),
    group = "Oregon Boundary"
  )

# Add year-wise layers
for (yr in 2009:2014) {
  layer_data <- oregon_race_merged %>%
    filter(year == yr)

  map <- map %>%
    addPolygons(
      data = layer_data,
      fillColor = ~pal(ell),
      color = "gray", weight = 0.5,  # Softer borders
      fillOpacity = 0.85, smoothFactor = 1,
      highlightOptions = highlightOptions(weight = 2, color = "black", bringToFront = TRUE),
      popup = ~paste0(
        "<b>", leaname, " (", yr, ")</b><br>",
        "Black Students: ", formatC(black, format = "d", big.mark = ","), "<br>",
        "White Students: ", formatC(white, format = "d", big.mark = ","), "<br>",
        "Hispanic Students: ", formatC(hispanic, format = "d", big.mark = ","), "<br>",
        "ELL Students: ", formatC(ell, format = "d", big.mark = ",")
      ),
      group = as.character(yr)  # Assign each layer to a year
    )
}

# Add layer control for years
map <- map %>%
  addLayersControl(
    baseGroups = as.character(2009:2014),
    overlayGroups = c("Oregon Boundary"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    pal = pal, values = oregon_race_merged$ell,
    title = "English Language Learners",
    position = "topright",
    labFormat = labelFormat(big.mark = ","),
    opacity = 1
  ) %>%
  setView(lng = -120.5542, lat = 43.8041, zoom = 6)  # Centered & fixed on Oregon

# Print the map
map

```













